Forecasting the 2024 U.S. Presidential Election

### 2024-11-04 paper/paper.qmd
Update paper.

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(readr)
library(here)
library(knitr)
library(kableExtra)
library(scales)
library(arrow)
library(bayesplot)
library(rstanarm)

data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
```

@@

# Data {#sec-data}
The data used in this paper was obtained from the FiveThirtyEight website [@Best2024], which aggregates polling data and other relevant indicators to track the U.S. election landscape. We accessed this data using custom web scraping tools and APIs provided by FiveThirtyEight. The dataset includes information on pollster, state, start and end dates of polls, sample size, population type (e.g., registered voters or likely voters), and party affiliation.

## Overview
Our analysis was conducted using R [@citeR] along with several supporting packages. We used the **tidyverse** for data manipulation [@tidyverse], **readr** for data import [@readr], and **here** for file management [@here].

Dynamic reporting was handled with **knitr** [@knitr], while **kableExtra** allowed for enhanced table formatting [@kableExtra]. We used **scales** to adjust visualization scales [@scales], and **arrow** for efficient data processing [@arrow].

For Bayesian modeling and visualization, we used **bayesplot** [@bayesplot] and **rstanarm** [@rstanarm].

## Measurement
The FiveThirtyEight dataset is updated in real-time, providing the latest polling data for the 2024 U.S. presidential election. This dataset aggregates information from multiple reputable polling sources, applying adjustments to enhance reliability, such as weighting based on pollster quality and recency of data. These measures ensure that the dataset remains a dependable reflection of current voter sentiment and trends as Election Day approaches.

## Analysis Data
The analysis dataset includes several key variables: State, Sample Size, Population Type (e.g., Registered Voters), Party Affiliation, Candidate Preference (Trump or Harris), and Pollster. These variables allow for an analysis of voter sentiment in specific swing states, with a focus on support levels between Donald Trump and Kamala Harris.

Certain variables, such as Question ID, Start Date, End Date, and Hypothetical Questions, were removed as they were not necessary for this analysis. Additionally, data on other candidates was excluded to concentrate solely on the Trump vs. Harris matchup, aligning with the focus on the most relevant contenders in the seven swing states: Pennsylvania, Georgia, North Carolina, Michigan, Arizona, Wisconsin, Nevada.

@tbl-sample provides a random sample of the analysis dataset, displaying all relevant variables used for analysis.

```{r}
#| label: tbl-sample
#| tbl-cap: Random sample of 10 rows from the analysis data.
#| echo: false
#| warning: false
#| message: false

set.seed(311)

# Sample 10 random observations from the filtered data with Trump and Harris, and remove unnecessary columns
sample_data <- data %>%
  filter(candidate_name %in% c("Donald Trump", "Kamala Harris")) %>%
  sample_n(10) %>%
  select(-start_date, -end_date, -question_id, -hypothetical, -pct)

# Display the sampled data as a table with refined styling
kable(sample_data, caption = "Sample of 10 Random Observations from the Analysis Data") %>%
  kable_styling(full_width = FALSE, position = "center") %>%  # Center the table in the PDF
  column_spec(1, width = "3cm") %>%    # Adjust specific columns for better fit
  column_spec(2, width = "2.5cm") %>%
  column_spec(3, extra_css = "padding-left: -5px;") %>%  # Move sample_size column slightly to the left
  column_spec(4:5, width = "1.5cm")    # Adjust remaining columns with less text

@@

## Swing State Polling Prediction Model
@fig-4 shows the trend of support rate changes for each candidate before the upcoming election, as well as the model's predicted values for the next 20 days.
```{r}
#| label: fig-4
#| fig-cap: "Swing State Polling Prediction Model"
#| echo: false
#| warning: false
#| message: false

invisible(Sys.setlocale("LC_TIME", "en_US"))
data_filtered <- read_parquet(here::here("data/02-analysis_data/model_data.parquet"))
model <- readRDS(here::here("models/model.rds"))
data_filtered <- data_filtered %>%
  mutate(predicted_pct = predict(model, newdata = ., type = "response"))

last_date <- max(data_filtered$end_date, na.rm = TRUE)
new_dates <- seq.Date(from = last_date + 1, by = "day", length.out = 20)
new_data <- expand.grid(
  end_date = new_dates,
  candidate_name = unique(data_filtered$candidate_name)
)
new_data <- new_data %>%
  mutate(
    end_date_numeric = as.numeric(end_date), 
  ) %>% mutate(predicted_pct = predict(model, newdata = ., type = "response"))
all_data <- bind_rows(data_filtered, new_data)

@@

## Influence of Endorsements and High-Profile Backing
Recently, high-profile figures like Elon Musk have publicly thrown their support behind Trump, leveraging their influence to rally support for his campaign [@Zahn2024]. Musk’s endorsement could invigorate Trump’s base, especially among Musk’s followers and those engaged in the fields of business and technology. This visible backing from such influential figures has the potential to sway undecided voters and influence last-minute decisions, particularly in swing states where Trump’s lead remains narrow.

Musk’s endorsement might not only energize Trump’s existing supporters but could also lead to a boost in campaign funds and social media engagement, further strengthening Trump’s position. In closely contested states, where a few thousand votes can make a significant difference, this level of backing from a high-profile figure could be crucial. If Musk’s involvement draws in additional resources and amplifies Trump’s visibility, it may ultimately impact the final outcome of the election on November 5th.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
posterior_predictions <- posterior_predict(model, draws = 100)
ppc_dens_overlay(y = data_filtered$average_pct/100, yrep = posterior_predictions) +
  labs(title = "Posterior Predictive Check",
       x = "Rate of support",
       y = "Density")
```

@@

>>>>

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(readr)
library(here)
library(knitr)
library(kableExtra)
library(scales)
library(arrow)
library(bayesplot)
library(rstanarm)
data_filtered <- read.csv(here::here("data/02-analysis_data/model_data.csv"))
data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
```

@@

# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

```{r}
#| echo: false
#| warning: false
#| message: false

# Set seed for reproducibility
set.seed(316)

# Sample 10 random observations from the filtered data with Trump and Harris
sample_data <- data %>%
  filter(candidate_name %in% c("Donald Trump", "Kamala Harris")) %>%
  sample_n(10)

# Display the sampled data as a table with smaller text and adjusted column spacing
kable(sample_data, caption = "Sample of 10 Random Observations from the Analysis Data") %>%
  kable_styling(font_size = 8, full_width = FALSE) %>%  # Set full_width to FALSE to center the table
  column_spec(1:ncol(sample_data), width = "2cm")  # Adjust width as needed, here set to 2cm per column

@@

## Swing State Polling Prediction Model
@fig-4 shows the trend of support rate changes for each candidate before the upcoming election, as well as the model's predicted values for the next 20 days.We predicts that Donald Trump may win based on our model.
```{r}
#| label: fig-4
#| fig-cap: "Swing State Polling Prediction Model"
#| echo: false
#| warning: false
#| message: false

invisible(Sys.setlocale("LC_TIME", "en_US"))
model_data <- read_csv(here::here("data/02-analysis_data/model_data.csv"))
model <- readRDS(here::here("models/model.rds"))

model_data <- model_data %>%
  mutate(predicted_pct = predict(model, newdata = ., type = "response"))
last_date <- max(model_data$end_date, na.rm = TRUE)
new_dates <- seq.Date(from = last_date + 1, by = "day", length.out = 20)
new_data <- expand.grid(
  end_date = new_dates,
  candidate_name = unique(model_data$candidate_name)
)
new_data <- new_data %>%
  mutate(
    end_date_numeric = as.numeric(end_date), 
  ) %>% mutate(predicted_pct = predict(model, newdata = ., type = "response"))
all_data <- bind_rows(model_data, new_data)

@@

## Influence of Endorsements and High-Profile Backing
Recently, high-profile figures like Elon Musk have publicly backed Trump, using their influence to mobilize support for his campaign. Musk’s endorsement may energize Trump’s voter base, especially among his followers and individuals interested in business and technology. This kind of visible support from influential figures could affect last-minute voter decisions, particularly in tightly contested states where Trump’s lead is slim. If Musk’s endorsement translates into increased campaign funding or social media influence, it could bolster Trump’s standing in swing states where the margin remains close, potentially altering the final outcome on November 5th.

@@

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows that the posterior predictive density closely follows the observed data density. Discrepancies between the two suggest that the model may be missing some aspect of the data structure or variability, which could guide potential model adjustments or refinement.

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows the prior for each parameter is centered around zero, but the posterior shows more precise estimates for each candidate and the date variable, reflecting the data's influence. Also, the posterior narrowing suggests that the model successfully extracted information from the data, reducing uncertainty around parameter estimates compared to the prior.
model_data <- read_parquet(here::here("data/02-analysis_data/model_data.parquet"))

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
pp_check(model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()

@@

@fig-stanareyouokay-1 is a trace plot. It shows chain mixing, stationarity and no drifting. This suggests that the model's MCMC sampling achieved convergence for each parameter. The chains are well-mixed and stationary, providing confidence in the posterior estimates derived from this sampling process.

@fig-stanareyouokay-2 is a Rhat plot. It shows the Rhat values for the parameters estimated in the model. Values close to 1 indicate that the chains have mixed well and the posterior distribution has been adequately explored. This suggests that the model has converged successfully and the estimates are reliable for inference.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(model, "trace")

plot(model, "rhat")
```
